\vspace{-1in} Demystifying the Computer Science \\PhD Admission in US Universities\\{\large A guide for International Students

### 2023-08-13 phd-cs-us.tex
Updates from Overleaf

Note that while having good grades at unknown school might not help,
having very bad grades will be \red{red flag} (unless your LORs or
statements give proper explanation). This is especially true if you
have bad grades in relevant, e.g., CS and Math, courses.

\begin{commentbox}[Thanh]
Vietnamese universities typically offer specialized programs, such as the talented engineer program at HUST, that have highly competitive entrance exams and a limited number of available slots (e.g., 30 per year). However, these programs often set higher requirements for students, including more demanding tests and assignments, resulting in lower GPAs and overall rankings. For example, an 3.5 GPA students from such talented programs are typically much better than a 4.0 GPA students not in those programs.  Similarly, variations in GPA standards exist among different universities, with technical universities generally having lower GPAs than economical universities. These make gaining admission in the US difficult as US faculty are not familiar with these issues.
\tcblower
\textbf{Vu}: Vietnamese students and even faculty often lament how this competitive/low-grading system hurts Vietnamese students when applying abroad. One way to mitigate this is making these issues known to admission committee in your statement.  Schools with Vietnamese profs are probably aware of them, but in general your letter writers/you can explicitly mention these in their letters/ your statement.
\end{commentbox}

@@

By around March you should hear back from most PhD programs you applied to . If you haven't heard back, reach our via email and ask about the status of your application.
If you receive offers, congratulations!  Now you're at a different game because the schools that have admitted you will try to get you to accept them!   You will likely have to make your decision by around April 15.

@@

As mentioned, if you're admitted to a \emph{good} CS PhD program, you should not have to worry about funding!  
In the US, the common types of funding for PhD are \emph{graduate teaching assistant} (GTA or TA), \emph{graduate research assistant} (GRA or RA), and \emph{Fellowship}.
RA is paid by a prof. for you to do their research. TA is paid by the department for you to help with teaching. Finally, fellowship is an independent funding that come from a school, a company, or an organization. Tab.~\ref{tab:funding} summarizes the differences.

Note that funding is typically more available for PhD students than 
Masters and especially undergraduate studies, which typically have no funding (and they also have to pay international tuition, which is very expensive!).

@@

Some words stipend. First, the amount of stipend depends on the university, which in turns depend on various factors such as location (e.g., a stipend in Washington DC is likely higher than in Lincoln, Nebraska due to higher living cost). Second, a school year is (typically) 9-month in the US, so stipend is for 9 months (so divide by 9 for each month). Third, like most source of income in the US, you will have to pay tax on your stipend.  Fourth, CS department typically has higher stipend comparing to other study fields.  Finally, private universities might pay more for stipend (but they could have higher "activity" or some other hidden fee, or you will have to pay some amount for each class credits).

Students often complain their stipend being too low, but it is actually not bad and you can live comfortably with it.  In many cases, it is enough to support your spouse and kids (many CS PhD students have their spouse and kids with them). So don't worry too much about stipend.  If you're admitted to a good CS PhD program, you will be fine.

@@

\begin{commentbox}[Vu]
TA and RA at GMU have similar benefits in tuition waiving and insurance.  The college and department will set the rate for 9-month graduate assistant stipend.  TA will likely be that amount but RA might be higher depending on the stage of the student (1st year vs ABD\footnote{All but dissertation: really close to graduate.}) and the prof. 
\tcblower
Having health insurance is required at many US universities.  So do not think that you're young and healthy and ignore insurance.  At GMU, and at most good CS PhD programs, your GTA or GRA \emph{will always} come with full insurance. In fact, at GMU your spouse/children will get significant discount rate for health insurance.  So you will never have to worry much about health issues for you or your family here.
\end{commentbox}

@@

TA is common in the beginning when you haven't found your advisor who would pay you RA. As a TA, you spend up to 20 hrs/week and help professors with their classes (e.g., grading or teaching labs/recitation). Your TA is paid through the school/department, i.e., they hire you to help teach.  During a semester, a TA might work with several courses and professors (not necessary their advisor).  TA funding typically is not available during the summer, which has no school.

@@

\begin{commentbox}[Vu]
If you got recruited by a prof. who would give you RA right away, it's very likely you will get admitted.  For example, if a prof., even if not in PhD admission committee, wants to work with and funds you, the PhD admission committee will respect that decision and admit your application (unless your application has many red flags).
\end{commentbox}

@@

\subsection{What can you do to increase your admission chance?}

 Show something that makes you \textbf{stand out}, e.g., do you have a degree or background in music and want to integrate that to CS? are you a female or a minority in CS (research for this on Google)? Do you participate in outreach activities that help increase diversity and inclusion in CS? All of these are unique and would get noticed from reviewers.

@@

You can learn about CS programs and research expertise of faculty using resources such as \href{https://csrankings.org}{CSRankings.org}, which is designed specifically to help prospective PhD students in Computer Science!  You will be very surprised to learn that a school that you didn't know much about has very strong research in your interested field (and vice versa, a school you thought highly off has no one works in the research field you're interested in). This is also a good way to learn about individual faculty (who works on what) and well-known CS conferences\footnote{In CS (and only in CS), conferences, not journals, are often the main venue to publish research finding.}. \S\ref{sec:ranking} gives the top 50 CS programs in the US according to CSRankings.

\begin{commentbox}[Dat] Most Vietnamese students, including those from top schools, do not know about CSRankings.  May be applicants who worked at top research places such as VinAI would know about it.
\end{commentbox}

More generally, rankings can be superficial and you need to do more research to be informed and make better decision. For example, if you get admissions to several places, you should contact profs. that you're interested in at those place and talk to them. They would be more willing to chat to you now that you have been admitted.  Ask them questions about their work, how they manage students, their expectations. You can even ask to contact their students.

@@

If you want to contact a prof. to ask about \emph{research opportunities}, or \emph{GTA/GRA} support, then \emph{yes}, I believe you should---it is \emph{worth it}. 
First, read the prof's website, see if they say something about contacting them. Many profs. indicate how prospective students should contact them (e.g., specific email subjects). 
In general, the best way to catch the prof.'s attention is to \emph{customize your email} to that faculty.  For example, read their papers, know what they work on, and see if you are interested in their research. Then send them an email talking how/why their work would match yours.
In contrast, if you write a generic email that can be sent to multiple professors (e.g., if you just change some names and keywords in the email), you will not likely get a response.

@@

Yes, as long as you can demonstrate you are ready for CS PhD research through research experiences, LoRs, statements, etc as mentioned. You might be even able to leverage this to make your profile stand out.

\subsection{Is an MS degree required for admission to PhD?}
No. In fact, student with BS can get MS degree ``along the way'' to PhD.  However, MS can help if it gives research experience or is from a more well-known school than your undergrad institution.

@@

It is worth it to note that if your intention is a PhD in CS, then you likely need to be in the CS dept \emph{and} advised by a CS faculty. For example, if your advisor is not in CS, then you likely cannot not get a CS degree. In fact, if a faculty is not in CS, they likely cannot be the main advisor of a CS PhD student---they may \emph{co-advise} or be in the PhD dissertation committee, but your main advisor will need to be a tenured or tenure-track faculty in CS. If in doubt, you should check with the CS department for their requirements.

Note that for the purpose of helping students interested in a PhD in CS, CSRankings only includes tenured or tenured track faculty who can advise CS PhD students. I also have compiled a \href{https://github.com/dynaroars/dynaroars.github.io/wiki/Viet-CS-Profs-US}{list} of Vietnamese faculty who can advise PhD students.

@@

\begin{wrapfigure}{l}{0.5\linewidth}
\vspace{-0.3in}
      \begin{center}
        \includegraphics[width=1.0\linewidth]{c5.png}
      \end{center}
      \vspace{-0.3in}
    \end{wrapfigure}
If you don't know the professor (e.g., first email contact), then use \textbf{Prof. Lastname} or \textbf{Dr. Lastname}. I've seen many international students write \textbf{Prof.} or \textbf{Dr.} \textbf{Firstname Lastname}.  Writing like that makes it like you copy and paste the names, so no need to do so,  just Prof. or Dr. Lastname.

If you don't know the prof., do not use Mr. or Mrs., or Firstname. To me it seems a bit disrespectful. As you know that prof. better and depends on their preference, you can call them by their Firstname.

>>>>

Note that while having good grades at unknown school might not help,
having very bad grades will be \red{red flag} (unless your LORs or
statements give proper explanation). This is especially true if you
have bad grades in relevant courses (e.g., Math and CS).

\begin{commentbox}[Thanh]
Vietnamese universities typically offer specialized programs, such as the talented engineer program at HUST, that have highly competitive entrance exams and a limited number of available slots (e.g., 30 per year). However, these programs often set higher requirements for students, including more demanding tests and assignments, resulting in lower GPAs and overall rankings. For example, an 3.5 GPA students from such talented programs are typically much better than a 4.0 GPA students not in those programs.  Similarly, variations in GPA standards exist among different universities, with technical universities generally having lower GPAs than economical universities. These make gaining admission in the US difficult as US faculty are not familiar with these issues.
\tcblower
\textbf{Vu}: Vietnamese students and even faculty often lament how this competitive/low-grading system hurts Vietnamese students when applying abroad. One way to mitigate this is making these issues known to admission committee in your statement.  Universities with Vietnamese profs are probably aware of them, but in general your letter writers \emph{and} you can explicitly mention these in their letters and your statement.
\end{commentbox}

@@

By around March you should hear back from most PhD programs you applied to . If you haven't heard back, reach out via email and ask about the status of your application.
If you receive offers, congratulations!  Now you're at a different game because the schools that have admitted you will try to get you to accept them!   You will have to make your decision by around April 15.

@@

As mentioned, if you're admitted to a \emph{good} CS PhD program, you should not have to worry about funding!  
In the US, the common types of funding for PhD are \emph{graduate teaching assistant} (GTA or TA), \emph{graduate research assistant} (GRA or RA), and \emph{Fellowship}.
RA is paid by a prof. for you to do their research. TA is paid by the department for you to help with teaching. Finally, fellowship is an independent funding that can come from a school, a company, or an organization. Tab.~\ref{tab:funding} summarizes the differences.

Note that funding is typically more available for PhD students than 
Masters. Undergraduate students typically have no funding and also have to pay international tuition, which is very expensive!.

@@

Several things about stipends.  First, the amount of stipend depends on the university, which in turns depend on various factors such as location (e.g., a stipend in Washington DC is likely higher than in Lincoln, Nebraska due to higher living cost). Second, a school year is (typically) 9-month in the US, so stipend is for 9 months (so divide by 9 for each month). Third, like most source of income in the US, you will have to pay tax on your stipend.  Fourth, CS department typically has higher stipend comparing to other study fields.  Finally, private universities might pay more for stipend (but they could have higher "activity" or some other hidden fee, or you will have to pay some fee for each class credits).

Students often complain their stipend being too low, but it is actually not bad and you can live comfortably with it.  In many cases, it is enough to support your spouse and kids (many CS PhD students have their spouse and kids with them). So don't worry too much about stipend.  If you're admitted to a good CS PhD program, you will be fine. A good school would know that it has to be competitive, otherwise it will loose good students to other schools.  For example, at GMU, every year we discuss about improving the benefits, and especially stipend, for our graduate students.

@@

\begin{commentbox}[Vu]
TA and RA at GMU have similar benefits in tuition waiving and insurance.  The college and department will set the rate for 9-month graduate assistant stipend.  TA, which is paid by the department, will likely be that amount but RA might be higher depending on the stage of the student (1st year vs ABD\footnote{All but dissertation: really close to graduate.}) and the prof. 
\tcblower
Having health insurance is required at many US universities.  So do not think that you're young and healthy and ignore insurance.  At GMU, and at most good CS PhD programs, your GTA or GRA \emph{will always} come with full insurance. In fact, at GMU your spouse/children will get significant discount rate for health insurance.  So you will never have to worry much about health issues for you or your family here.
\end{commentbox}

@@

TA is common in the beginning when you haven't found your advisor who would pay you RA. As a TA, you spend up to 20 hrs/week and help professors with their classes (e.g., grading or teaching labs/recitation). Your TA is paid through the school/department, i.e., they hire you to help teach.  During a semester, a TA might work with several courses and professors (not necessary their advisor).  TA funding typically is not available during the summer, which has much fewer courses.

@@

\begin{commentbox}[Vu]
If you got recruited by a prof. who would give you RA right away, it's very likely you will get admitted.  For example, if a prof., even if not in PhD admission committee, wants to work with and funds you, the PhD admission committee will respect that decision and admit you (unless your application has many red flags).
\end{commentbox}

@@

\subsection{What can you do to increase your admission chance?}\label{sec:improve-your-chance}

 Show something that makes you \textbf{stand out}, e.g., do you have a degree or background in music and want to integrate that with CS? are you a female or a minority in CS (research for this on Google)? Do you participate in outreach activities that help increase diversity and inclusion in CS? All of these are unique and would get noticed from reviewers.

@@

You can learn about CS programs and research expertise of faculty using resources such as \href{https://csrankings.org}{CSRankings.org}, which is designed specifically to help prospective PhD students in Computer Science!  You will be very surprised to learn that a school that you didn't know much about can have very strong research in your interested topic (and vice versa, a school you thought highly about has no one works in the research field you're interested in). This is also a good way to learn about individual faculty (who works on what) and well-known CS conferences\footnote{In CS (and only in CS), conferences, not journals, are often the main venue to publish research finding.}. \S\ref{sec:ranking} gives the top 50 CS programs in the US according to CSRankings.

\begin{commentbox}[Dat] Most Vietnamese students, including those from top schools, \textbf{do not know} about CSRankings.  May be applicants who worked at top research places such as VinAI would know about it.
\end{commentbox}

However, in general, rankings can be superficial and you need to do more research to be informed and make better decision. For example, if you get admissions to several places, you should contact profs. that you're interested in at those place and talk to them. They would be more willing to chat to you now that you have been admitted.  Ask them questions about their work, how they manage students, their expectations. You can even ask to contact their students.

@@

If you want to contact a prof. to ask about \emph{research opportunities}, or \emph{GTA/GRA} support, then \emph{yes}, I believe you should---it is \emph{worth it}. However, you need to do it in a right way.

First, read the prof's website, see if they say something about contacting them. Many profs. indicate how prospective students should contact them (e.g., specific email subjects). 
In general, the best way to catch the prof.'s attention is to \emph{customize your email} to that faculty.  For example, read their papers, know what they work on, and see if you are interested in their research. Then send them an email talking how/why their work would match yours.
In contrast, if you write a generic email that can be sent to multiple professors (e.g., if you just change some names and keywords in the email), you will not get a response.

@@

Yes, as long as you can demonstrate you are ready for CS PhD research through research experiences, LoRs, statements, etc as mentioned. You might be even able to leverage this to make your profile stand out as mentioned in \S\ref{sec:improve-your-chance}.

\subsection{Is an MS degree required for admission to PhD in CS?}
No. In fact, student with BS can get MS degree ``along the way'' to PhD.  However, MS can help if it gives research experience or is from a more well-known school than your undergrad institution.

@@

It is worth it to note that if your intention is a PhD in CS, then you likely need to be in the CS dept \emph{and} advised by a CS faculty. For example, if your advisor is not in CS, then you likely cannot not get a CS degree. In fact, if a faculty is not in CS, they likely cannot be the main advisor of a CS PhD student---they may \emph{co-advise} or be in the PhD dissertation committee, but your main advisor will need to be a tenured or tenure-track faculty in CS. If in doubt, you should check with the CS department for their requirements.  For this specific reason,  CSRankings only includes tenured or tenured track faculty who can advise CS PhD students. I also have compiled a \href{https://github.com/dynaroars/dynaroars.github.io/wiki/Viet-CS-Profs-US}{list} of Vietnamese faculty who can advise PhD students.

@@

\begin{wrapfigure}{l}{0.5\linewidth}
\vspace{-0.3in}
      \begin{center}
        \includegraphics[width=1.0\linewidth]{c5.png}
      \end{center}
      \vspace{-0.5in}
    \end{wrapfigure}
If you don't know the professor (e.g., first email contact), then use \textbf{Prof. Lastname} or \textbf{Dr. Lastname}. I've seen many international students write \textbf{Prof.} or \textbf{Dr.} \textbf{Firstname Lastname}.  Writing like that makes it like you copy and paste names, so no need to do so,  just Prof. or Dr. Lastname.

If you don't know the prof., do not use Mr. or Mrs., or Firstname. To me it seems a bit disrespectful. As you know that prof. better and depends on their preference, you may call them by their Firstname.

Note that at some universities, especially those that do not offer PhD studies, will use the formal \textbf{Dr.} title, i.e., Dr. Lastname.  So just observe and follow how it is done at your place.
